One-dimensional Classical Gas in a Harmonic Trap with Repulsive Interaction

### 2017-12-08 draft_05.tex
working on thermalization section

\section{Thermalization}\label{section:Thermalization}
\subsection{Dynamics Study}
It is natural to think that a many-body system with interaction will be thermalized in ``usual" case. To study the condition of thermalization is equivalent to find out the mechanism that prohibits the system from ergodic. We begin the study with two-particle motion. 
\subsubsection{two particles}

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{ZhiyuPictures/effective_potential.png}
\caption{effective potential for two-particle system}
\label{fig:thermalization2}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

To see the way interaction change the orbit (energy level) of the particles, we had better start from two-particle case.

We can consider their motion in their center of mass frame ``C". In frame C, one can easily show that when there is no interaction what we will see is that two particles bound together by a harmonic trap centered at C (black line in Fig.\ref{fig:thermalization2}). So their relative motion is also a harmonic oscillation.

In this case, the system has two frequency components. The first one is the frequency of C, which is just $\omega_0$ . The other one is the frequency of their relative motion $\omega_r$, which is slightly deviated from $\omega_0$. The deviation grows with the increase of $F_0$ and the decrease of internal energy of the pair. As a result, when $F_0$ is small or when E is large, a beat with low frequency $\omega_r-\omega_0=\delta$ will exist. This phenomenon means energy transfer between particles are inefficient, thus the thermalization should be slow.

\begin{figure}[h]

\subfigure[$E_i=2$]{
\begin{minipage}[b]{0.22\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=2_single_particle_energy_600_500_Font24.eps} 
%\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=2_phasespace.eps}
\end{minipage}
}
\subfigure[$E_i=3$]{
\begin{minipage}[b]{0.22\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=3_single_particle_energy_600_500_Font24.eps} 
%\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=3_phasespace.eps}
\end{minipage}
}
\subfigure[$E_i=5$]{
\begin{minipage}[b]{0.22\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=5_single_particle_energy_600_500_Font24.eps} 
%\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=5_phasespace.eps}
\end{minipage}
}
\subfigure[$E_i=10$]{
\begin{minipage}[b]{0.22\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=10_single_particle_energy_600_500_Font24.eps} 
%\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=10_phasespace.eps}
\end{minipage}
}

\caption{the dependence of beat frequency on internal energy. The diagram is the energy of two
  particles. Here, $F_0$ is set to be 1. In this case, the larger the $E_i$, the lower the beat
  frequency. \label{fig:thermalization1}}
%Actually in very low energy, we can retrieve a beat, but that is for different reason -- it is a
%beat between $\omega_0$ and $\frac{\omega_0}{2}+\delta $

\end{figure}

\subsubsection{More particles}
Now let's consider the three-particle case. For two particle, the motion is non-chaotic. On the other hand, intuitively, we will say that three-body motion is chaotic so that the system could be ``thermalized" soon. However, in some cases, the time scale of thermalizing could still be very long. Suppose two of them, say, A and B, has small internal energy, which means their distance and relative velocity are both small. Meanwhile, suppose particle C has some energy quite different from A \& B. In this case, A \& B will often be in the interaction, while C will pass both of them at a high speed in each period. How will energy transfer between C and the two-particle system A and B? Since the relative velocity of C and the two-particle system is usually large, C will pass A-B pair in a short time $\tau\sim\frac{\sigma}{R}$ ($\tau\ll\frac{2\pi}{\omega}$). C gives A a push when approaching it, and then push A back when leaving.  As has been discussed in the two-particle case, this process is equivalent to giving A a very small velocity ($\sim O(\frac{\sigma}{R})$) in the background of a trap. Since both the position and velocity of A and B are close($|x_A-x_B|\sim\sigma$), the velocity increase of A \& B are almost the same (difference$\sim O(\frac{\sigma^2}{R^2})$). In this manner, the passing of particle C only kicks the center of mass of the A-B pair slightly, leaving the internal motion of the pair unaffected. In another word, the existence of C will not have significant effect on the energy transfer between A and B, but only ``dance" with their center of mass. The physics of the ``dance" is similar to the dance between two particle. As is shown in Fig.\ref{fig:thermalization4}, the particles with energy close to each other tends to form a pair with small internal energy and the pair's internal energy transfer is relatively stable.

>>>>

\section{Thermalization}\label{section:Thermalization}

For any nonzero interaction, the system is expected to be ergodic for $N>2$.  Once there are many
particles, one expects thermalization in the long-time limit.  From the few-particle perspective,
several interesting questions pose themselves.  First, although we expect ergodicity, the question
of how long it takes to thermalize is an open question for small $N$. We treat below a
coarse-grained version of this question: namely, we ask whether particles show ergodic behavior
within reasonable timescales chosen to be (somewhat arbitrarily) in the timestep range of
$10^3$--$10^4$.  Another question is the connection between thermalization, as defined by the
appearance of a Boltzmann distribution and other intuitive characteristics of ergodic behavior such
as whether the single-particle phase space is isotropically occupied.  We find cases where one
aspect is seen while the other is not.

In the subsections below, we first ***  then ****

\subsection{Few-particle considerations}

Since we are interested in thermalization within finite timescales, it is useful to consider
mechanisms which hinder relaxation or thermalization.  We begin with considerations of few-particle
motion.  Since ergodicity and relaxation are generally expected to be more robust and efficient with
larger particle numbers, consideration of small particle numbers will highlight effects which slow
down relaxation.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbph]
\centering
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/two_particles_a_01.pdf}
\caption{Left: Effective potential for two-particle system.  Center and right: beats at internal
  energy $E_i=2$ and $E_i=5$.  The curves show the individual energies of the two particles.  Here
  $F_0=1$. In this case, the larger the $E_i$, the lower the beat frequency. }
\label{fig:thermalization2}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

\subsubsection{Two particles}

We consider the two-particle motion in their center of mass frame.  The center of mass itself
executes simple harmonic oscillation.  In the absence of interactions ($F_0=0$), the effective
potential within the center of mass frame is itself parabolic, with the same frequency.  So their
relative motion is also a harmonic oscillation, with the same frequency $\omega_0$.

With interactions, a term $F_0(1-|r|)\theta(1-|r|)$ is added, where $r$ is the relative
displacement; as shown in Figure \ref{fig:thermalization2}.  Now, the relative motion is no longer
harmonic; one could regard the resulting motion has having a slightly different frequency, or a
collection of frequencies whose center is shifted slightly from $\omega_0$.  Since the
center-of-mass motion is still of frequency $\omega_0$, we have a superposition of slightly
different frequencies, resulting in beating dynamics.  This is clearly seen in the time evolution of
individual energies shown in Figure \ref{fig:thermalization2}.  The two panels correspond to
different internal energy $E_i$ (defined as the total energy minus the center-of-mass energy).  For
larger $E_i$, the distortion of the effective potential (at constant $F_0$) plays a smaller role in
shifting the effective frequency of relative motion; hence the beat frequency is smaller.

This illustrates a simple mechanism hindering the redistribution of energy between particles, which
is necessary for thermalization or relaxation.  As seen in the example dynamics shown, the
difference in energy between the two particles is sustained over time.  Of course thermalization is
not expected anyway in a two-particle system, but we will see below how this basic effect continues
to play a role for larger $N$.

-------

%% \begin{figure}[h]
%% \subfigure[$E_i=2$]{
%% \begin{minipage}[b]{0.22\textwidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=2_single_particle_energy_600_500_Font24.eps} 
%% %\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=2_phasespace.eps}
%% \end{minipage}
%% }
%% \subfigure[$E_i=3$]{
%% \begin{minipage}[b]{0.22\textwidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=3_single_particle_energy_600_500_Font24.eps} 
%% %\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=3_phasespace.eps}
%% \end{minipage}
%% }
%% \subfigure[$E_i=5$]{
%% \begin{minipage}[b]{0.22\textwidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=5_single_particle_energy_600_500_Font24.eps} 
%% %\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=5_phasespace.eps}
%% \end{minipage}
%% }
%% \subfigure[$E_i=10$]{
%% \begin{minipage}[b]{0.22\textwidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=10_single_particle_energy_600_500_Font24.eps} 
%% %\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=10_phasespace.eps}
%% \end{minipage}
%% }

%% \caption{the dependence of beat frequency on internal energy. The diagram is the energy of two
%%   particles. Here, $F_0$ is set to be 1. In this case, the larger the $E_i$, the lower the beat
%%   frequency. \label{fig:thermalization1}}
%% %Actually in very low energy, we can retrieve a beat, but that is for different reason -- it is a
%% %beat between $\omega_0$ and $\frac{\omega_0}{2}+\delta $
%% %
%% \end{figure}

\subsubsection{More particles}

Now let's consider the three-particle case. For two particle, the motion is non-chaotic. On the other hand, intuitively, we will say that three-body motion is chaotic so that the system could be ``thermalized" soon. However, in some cases, the time scale of thermalizing could still be very long. Suppose two of them, say, A and B, has small internal energy, which means their distance and relative velocity are both small. Meanwhile, suppose particle C has some energy quite different from A \& B. In this case, A \& B will often be in the interaction, while C will pass both of them at a high speed in each period. How will energy transfer between C and the two-particle system A and B? Since the relative velocity of C and the two-particle system is usually large, C will pass A-B pair in a short time $\tau\sim\frac{\sigma}{R}$ ($\tau\ll\frac{2\pi}{\omega}$). C gives A a push when approaching it, and then push A back when leaving.  As has been discussed in the two-particle case, this process is equivalent to giving A a very small velocity ($\sim O(\frac{\sigma}{R})$) in the background of a trap. Since both the position and velocity of A and B are close($|x_A-x_B|\sim\sigma$), the velocity increase of A \& B are almost the same (difference$\sim O(\frac{\sigma^2}{R^2})$). In this manner, the passing of particle C only kicks the center of mass of the A-B pair slightly, leaving the internal motion of the pair unaffected. In another word, the existence of C will not have significant effect on the energy transfer between A and B, but only ``dance" with their center of mass. The physics of the ``dance" is similar to the dance between two particle. As is shown in Fig.\ref{fig:thermalization4}, the particles with energy close to each other tends to form a pair with small internal energy and the pair's internal energy transfer is relatively stable.
